The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer: A SEER-Based Cohort Study

### 2025-04-03 code/analysis-code/statistical-analysis.qmd
part 4 project

This Quarto document presents the statistical data analysis for this project, focusing on modeling. It includes bivariate analyses to identify associations between key variables and a multivariate Cox proportional hazards model to assess survival outcomes. The analysis is conducted using the tidymodels framework and survival package in R.
# Loading Package
```{r}
#load needed packages. make sure they are installed.
# Load required packages. Ensure they are installed.
library(survival)  # For Cox regression analysis
library(survminer) # For survival curve visualization
library(broom)     # For tidying model output
library(tidyverse) # For data manipulation
library(here)      # For file path management (ensures reproducibility)
library(stringr)  # Ensure string manipulation functions are available
```
# Loadinf data
```{r}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")

#load data. 
mydata <- readRDS(data_location)
```
# Bivariate Analysis
The dataset is preprocessed to ensure consistency for survival analysis. First, all "Unknown" values in metastasis variables are removed, retaining only "Yes" and "No" categories. Next, metastasis variables are converted to factors with "No" as the reference category to ensure proper hazard ratio interpretation in Cox regression. A survival object is then created using "Survival months" as the time variable and "Vital status recode (study cutoff used)" to define event occurrence. These steps standardize the data for valid and reproducible survival analysis.

@@

A Cox proportional hazards model is fitted separately for each metastasis variable to assess its impact on survival. The function dynamically constructs the regression formula and calculates hazard ratios (HRs) for "Yes" vs. "No" metastasis status. Results from all four models (bone, brain, liver, and lung metastases) are combined into a single table for better interpretation. The metastasis variable names are cleaned for clarity, and the final results are displayed and saved for reproducibility. This ensures a structured and systematic approach to analyzing the association between metastasis and survival outcomes.
## Bivariate Analysis: Metastasis and Survival
```{r}
####################################################
# Bivariate Analysis: Metastasis and Survival
####################################################
# Function to fit Cox regression for each metastasis variable
run_cox <- function(var) {
  formula <- as.formula(paste("surv_obj ~", paste0("`", var, "`")))  # Construct formula
  cox_model <- coxph(formula, data = mydata_filtered)  # Fit Cox model
  result <- broom::tidy(cox_model, exponentiate = TRUE)  # Convert log(HR) to HR
  return(result)  # Return model results
}

@@

```
## KM curve Metastasis Bone
```{r}
 #Convert vital_status to a binary variable
mydata_filtered$survival_status <- ifelse(mydata_filtered$vital_status == "Dead", 1, 0)

@@

```
## KM curve Metastasis Brain
```{r}
 ####################################################
# Kaplan-Meier Analysis: Brain Metastasis and Survival
####################################################

@@

```
## KM curve Metastasis Liver
```{r}
####################################################
# Kaplan-Meier Analysis: Liver Metastasis and Survival
####################################################

@@

## KM curve Metastasis Lung 
```{r}
####################################################
# Kaplan-Meier Analysis: Lung Metastasis and Survival
####################################################

@@

The race variable is recoded as a factor with "White" as the reference category to ensure proper comparison in Cox regression. A separate model is fitted to assess the association between race and survival, with hazard ratios (HRs) calculated for Black, Asian or Pacific Islander, and American Indian/Alaska Native populations relative to White. The results are formatted for clarity, renaming the terms to indicate direct comparisons. Finally, the updated results are displayed and saved for reproducibility, allowing for structured and interpretable survival analysis by race.
```{r}
####################################################
# Bivariate Analysis: Race and Survival
####################################################

@@

## Bivariate Analysis: Age and Survival
The age variable is recoded as a factor with "20-24 years" set as the reference category to allow for meaningful comparisons in Cox regression. A separate model is fitted to evaluate the association between different age groups and survival outcomes, with hazard ratios (HRs) estimated relative to the 20-24 years category. The results are formatted to improve clarity, renaming each term to reflect direct age group comparisons. Finally, the updated results are displayed and saved for reproducibility, ensuring a structured and interpretable analysis of age-related survival differences.

@@

```
## Bivariate Analysis: Year of Diagnosis and Survival

@@

The median household income variable, inflation-adjusted to 2019, is recoded as a factor with "$75,000+" as the reference category to allow for clear comparisons in Cox regression. A separate model is fitted to assess the association between income levels and survival outcomes, with hazard ratios (HRs) estimated relative to the highest income group. The results are formatted for better readability by appending comparison labels for each income group against "$75,000+." Finally, the updated results are displayed and saved for reproducibility, ensuring a structured and interpretable analysis of socioeconomic disparities in survival.

@@

# Multivariable Cox Proportional Hazards Model
```{r}
####################################################
# Multivariable Cox Proportional Hazards Model
####################################################

@@

>>>>

This Quarto document presents the statistical data analysis for this project, focusing on modeling. It includes bivariate analyses to identify associations between key variables and a multivariate Cox proportional hazards model to assess survival outcomes. The analysis is conducted using the tidymodels framework and survival package in R. \# Loading Package

```{r}
#load needed packages. make sure they are installed.
# Load required packages. Ensure they are installed.
library(survival)  # For Cox regression analysis
library(survminer) # For survival curve visualization
library(broom)     # For tidying model output
library(tidyverse) # For data manipulation
library(here)      # For file path management (ensures reproducibility)
library(stringr)  # Ensure string manipulation functions are available
library(tidymodels)
library(vip)
library(themis)
library(randomForestSRC)
library(dplyr)
library(ranger)
library(survival)
library(pROC)
library(caret)
library(glmnet)
```

# Loading data

```{r}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")

#load data. 
mydata <- readRDS(data_location)
```

# Bivariate Analysis

The dataset is preprocessed to ensure consistency for survival analysis. First, all "Unknown" values in metastasis variables are removed, retaining only "Yes" and "No" categories. Next, metastasis variables are converted to factors with "No" as the reference category to ensure proper hazard ratio interpretation in Cox regression. A survival object is then created using "Survival months" as the time variable and "Vital status recode (study cutoff used)" to define event occurrence. These steps standardize the data for valid and reproducible survival analysis.

@@

A Cox proportional hazards model is fitted separately for each metastasis variable to assess its impact on survival. The function dynamically constructs the regression formula and calculates hazard ratios (HRs) for "Yes" vs. "No" metastasis status. Results from all four models (bone, brain, liver, and lung metastases) are combined into a single table for better interpretation. The metastasis variable names are cleaned for clarity, and the final results are displayed and saved for reproducibility. This ensures a structured and systematic approach to analyzing the association between metastasis and survival outcomes. \## Bivariate Analysis: Metastasis and Survival

```{r}
####################################################
# Bivariate Analysis: Metastasis and Survival
####################################################
# Function to fit Cox regression for each metastasis variable
run_cox <- function(var) {
  formula <- as.formula(paste("surv_obj ~", paste0("`", var, "`")))  # Construct formula
  cox_model <- coxph(formula, data = mydata_filtered)  # Fit Cox model
  result <- broom::tidy(cox_model, exponentiate = TRUE)  # Convert log(HR) to HR
  return(result)  # Return model results
}

@@

## KM curve Metastasis Bone

```{r}
 #Convert vital_status to a binary variable
mydata_filtered$survival_status <- ifelse(mydata_filtered$vital_status == "Dead", 1, 0)

@@

## KM curve Metastasis Brain

```{r}
 ####################################################
# Kaplan-Meier Analysis: Brain Metastasis and Survival
####################################################

@@

## KM curve Metastasis Liver

```{r}
####################################################
# Kaplan-Meier Analysis: Liver Metastasis and Survival
####################################################

@@

## KM curve Metastasis Lung

```{r}
####################################################
# Kaplan-Meier Analysis: Lung Metastasis and Survival
####################################################

@@

The race variable is recoded as a factor with "White" as the reference category to ensure proper comparison in Cox regression. A separate model is fitted to assess the association between race and survival, with hazard ratios (HRs) calculated for Black, Asian or Pacific Islander, and American Indian/Alaska Native populations relative to White. The results are formatted for clarity, renaming the terms to indicate direct comparisons. Finally, the updated results are displayed and saved for reproducibility, allowing for structured and interpretable survival analysis by race.

```{r}
####################################################
# Bivariate Analysis: Race and Survival
####################################################

@@

## Bivariate Analysis: Age and Survival

The age variable is recoded as a factor with "20-24 years" set as the reference category to allow for meaningful comparisons in Cox regression. A separate model is fitted to evaluate the association between different age groups and survival outcomes, with hazard ratios (HRs) estimated relative to the 20-24 years category. The results are formatted to improve clarity, renaming each term to reflect direct age group comparisons. Finally, the updated results are displayed and saved for reproducibility, ensuring a structured and interpretable analysis of age-related survival differences.

@@

## Bivariate Analysis: Year of Diagnosis and Survival

@@

The median household income variable, inflation-adjusted to 2019, is recoded as a factor with "\$75,000+" as the reference category to allow for clear comparisons in Cox regression. A separate model is fitted to assess the association between income levels and survival outcomes, with hazard ratios (HRs) estimated relative to the highest income group. The results are formatted for better readability by appending comparison labels for each income group against "\$75,000+." Finally, the updated results are displayed and saved for reproducibility, ensuring a structured and interpretable analysis of socioeconomic disparities in survival.

@@

# Multivariable Cox Proportional Hazards Model

```{r}
####################################################
# Multivariable Cox Proportional Hazards Model
####################################################

@@

# Machine Learning Model

I used the random forest for predicting cause specifc death but the model did not perform well. Therefore, using the random survival forest model.

A random forest classification model was developed to predict cause-specific death using demographic and clinical features, including metastasis status, age group, race, income, and year of diagnosis. The model was trained on 80% of the data and evaluated on a 20% test set, with hyperparameter tuning via 5-fold cross-validation. Despite identifying several important predictors—most notably lung and liver metastases, age group 45–49, bone metastasis, and race—the model demonstrated poor overall predictive performance. The area under the ROC curve (AUC) was 0.39, indicating that the model performed only marginally better than random chance in distinguishing between patients who died and those who survived. The confusion matrix further reflected this limitation, showing moderate sensitivity but poor specificity. These results suggest that while random forest can highlight relevant variables, it is not well-suited for modeling survival outcomes in the presence of censoring. In contrast, approaches such as random survival forests (RSF) or Cox-based models offer more reliable and interpretable performance in survival analysis contexts. \## Random Forest

```{r}
####################################################
# Random Forest with Class Balancing (Downsampling)
# Goal: Predict cause-specific death using SEER variables
####################################################

# Step 1: Prepare the dataset
rf_data <- mydata_filtered %>%
  select(cause_specific_death, age_group, race, median_income, 
         mets_bone, mets_brain, mets_liver, mets_lung, diagnosis_year) %>%
  drop_na()

# Step 2: Recode outcome to binary factor ("Yes"/"No")
rf_data <- rf_data %>%
  mutate(cause_specific_death = case_when(
    cause_specific_death == "Dead (attributable to this cancer dx)" ~ "Yes",
    cause_specific_death == "Alive or dead of other cause" ~ "No",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(cause_specific_death)) %>%
  mutate(cause_specific_death = factor(cause_specific_death, levels = c("No", "Yes")))

# Step 3: Split the data (80% training / 20% testing), stratified by class
set.seed(123)
rf_split <- initial_split(rf_data, prop = 0.8, strata = cause_specific_death)
rf_train <- training(rf_split)
rf_test  <- testing(rf_split)

# Step 4: Preprocessing recipe with downsampling for class balance
rf_recipe <- recipe(cause_specific_death ~ ., data = rf_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_downsample(cause_specific_death)

# Step 5: Define random forest model (with hyperparameter tuning)
rf_model <- rand_forest(mtry = tune(), trees = 500, min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# Step 6: Create workflow
rf_workflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_model)

# Step 7: 5-fold cross-validation
set.seed(123)
rf_folds <- vfold_cv(rf_train, v = 5)

# Step 8: Tune the model
rf_tuned <- tune_grid(
  rf_workflow,
  resamples = rf_folds,
  grid = 10,
  metrics = metric_set(roc_auc, accuracy)
)

# Step 9: Select best model based on ROC AUC
best_rf <- select_best(rf_tuned, metric = "roc_auc")

# Step 10: Finalize and train final model
final_rf <- finalize_workflow(rf_workflow, best_rf)
rf_fit <- fit(final_rf, data = rf_train)

# Step 11: Make predictions on test set
rf_predictions <- predict(rf_fit, rf_test, type = "prob") %>%
  bind_cols(predict(rf_fit, rf_test)) %>%
  bind_cols(rf_test)

# Step 12: Evaluate model
rf_auc <- roc_auc(rf_predictions, truth = cause_specific_death, .pred_Yes)
rf_cm <- conf_mat(rf_predictions, truth = cause_specific_death, estimate = .pred_class)

# Step 13: Variable importance plot
vip_plot <- rf_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 10)

# Step 14: Print results
rf_auc
rf_cm
vip_plot

# Step 15: Save outputs
saveRDS(rf_fit, here("results", "tables", "random_forest_balanced_model.rds"))
saveRDS(rf_auc, here("results", "tables", "random_forest_balanced_auc.rds"))
saveRDS(rf_cm, here("results", "tables", "random_forest_balanced_confusion_matrix.rds"))
ggsave(filename = here("results", "figures", "vip_plot_rf_balanced.png"), plot = vip_plot, dpi = 300, width = 8, height = 6)
```

## Random Survival Forest

A random survival forest model was trained on clinical and demographic features to predict survival outcomes. Variable importance analysis revealed that age group, median household income, diagnosis year, and race were the top contributors to survival prediction, whereas metastasis variables (bone, lung, liver, brain) were less influential. The model demonstrated moderate performance, with an AUC of \~0.72 at 12 months. While sensitivity was high (86.5%), specificity remained modest (47.9%), indicating that the model more accurately identified patients who survived. Individual-level survival curves were generated, illustrating patient-specific risk trajectories.

```{r}
####################################################
# Random Survival Forest (RSF)
# Goal: Model survival time using censored data
####################################################

# Step 1: Prepare and clean data
rsf_data <- mydata_filtered %>%
  mutate(event = vital_status) %>%
  select(survival_months, event,
         mets_lung, mets_liver, mets_bone, mets_brain,
         race, age_group, median_income, diagnosis_year) %>%
  # Drop rows with any NA in predictors or outcome
  filter(if_all(everything(), ~ !is.na(.))) %>%
  mutate(
    mets_lung = as.factor(mets_lung),
    mets_liver = as.factor(mets_liver),
    mets_bone = as.factor(mets_bone),
    mets_brain = as.factor(mets_brain),
    race = as.factor(race),
    age_group = as.factor(age_group),
    median_income = as.factor(median_income),
    diagnosis_year = as.factor(diagnosis_year)
  )

# Step 2: Create survival object
rsf_data$surv_obj <- with(rsf_data, Surv(survival_months, event))

# Step 3: Fit survival random forest using ranger
rsf_model <- ranger(
  formula = surv_obj ~ mets_lung + mets_liver + mets_bone + mets_brain +
    race + age_group + median_income + diagnosis_year,
  data = rsf_data,
  num.trees = 300,
  mtry = 3,
  splitrule = "logrank",  # For survival
  importance = "impurity",
  seed = 123
)

# Step 4: Plot variable importance
vip(rsf_model)

```{r}
####################################################
# RSF Results: Top Variables, Survival Prediction, AUC & Confusion Matrix
####################################################
# --- Step 1: Top Variable Importance ---
importance_df <- data.frame(
  variable = names(rsf_model$variable.importance),
  importance = rsf_model$variable.importance
) %>%
  arrange(desc(importance))

# View top 10
head(importance_df, 10)

# Save top 10
write.csv(head(importance_df, 10), here::here("results", "tables", "top_10_rsf_variables.csv"), row.names = FALSE)

# Plot top 10 variables
vip::vip(rsf_model, num_features = 10)

# --- Step 2: Predict Survival Curves ---
# Predict full survival curve for all patients
surv_preds <- predict(rsf_model, data = rsf_data, type = "response")

# Plot survival curve for the first patient
plot(rsf_model$unique.death.times, surv_preds$survival[1, ], type = "l",
     xlab = "Time (months)", ylab = "Survival Probability",
     main = "Survival Curve for Patient 1")

# --- Step 3: Classify Patients Based on Survival at 12 Months ---
# Choose cutoff time point (e.g., 12 months)
target_time <- 12
closest_index <- which.min(abs(rsf_model$unique.death.times - target_time))

# Predicted survival prob at 12 months
pred_surv_12 <- surv_preds$survival[, closest_index]

# Convert to binary prediction: 1 = died before 12 months (low survival prob)
pred_class <- ifelse(pred_surv_12 < 0.5, 1, 0)
true_class <- rsf_data$event  # Actual: 1 = dead, 0 = alive

# --- Step 4: Confusion Matrix ---
conf_matrix <- caret::confusionMatrix(factor(pred_class), factor(true_class))
print(conf_matrix)

# --- Step 5: ROC Curve and AUC ---
roc_obj <- roc(response = true_class, predictor = 1 - pred_surv_12)
auc_value <- auc(roc_obj)
print(paste("AUC at 12 months:", round(auc_value, 3)))

# Plot ROC
plot(roc_obj, main = "ROC Curve at 12 Months", col = "blue", lwd = 2)

# --- Step 6: Save Outputs ---
# Create and save top variable importance plot
importance_plot <- vip::vip(rsf_model, num_features = 10)

ggsave(
  filename = here::here("results", "figures", "rsf_variable_importance.png"),
  plot = importance_plot,
  width = 8,
  height = 6,
  dpi = 300
)

# Save survival curve plot using base R
png(filename = here::here("results", "figures", "rsf_survival_curve_patient1.png"),
    width = 800, height = 600)

plot(rsf_model$unique.death.times, surv_preds$survival[1, ], type = "l",
     xlab = "Time (months)", ylab = "Survival Probability",
     main = "Survival Curve for Patient 1", col = "darkgreen", lwd = 2)

dev.off()

# Save ROC curve plot using base R
png(filename = here::here("results", "figures", "rsf_auc_12mo.png"),
    width = 800, height = 600)

plot(roc_obj, main = "ROC Curve at 12 Months", col = "blue", lwd = 2)
dev.off()

# Save confusion matrix text
sink(file = here::here("results", "tables", "rsf_confusion_matrix.txt"))
print(conf_matrix)
sink()

# Optional: Save a visual confusion matrix (if you have `autoplot` or ggplot-based CM)
# e.g., using fourfoldplot for base R
png(filename = here::here("results", "figures", "rsf_conf_matrix_12mo.png"),
    width = 800, height = 600)

fourfoldplot(conf_matrix$table, color = c("#E41A1C", "#377EB8"),
             main = "RSF Confusion Matrix (12 Months)")

dev.off()

## Lasso-Cox Model

A LASSO-penalized Cox proportional hazards model was applied to the survival dataset to perform variable selection and reduce model complexity. Cross-validation identified the optimal penalty parameter (λ) that minimized the partial likelihood deviance. The model selected 10 predictors with non-zero coefficients, including all four metastasis variables (mets_lung, mets_liver, mets_bone, mets_brain) and several younger age groups (with negative coefficients, indicating lower risk). The results support the prognostic value of metastasis presence and younger age in overall survival prediction.

```{r}
####################################################
# LASSO Cox Regression using glmnet
####################################################
# Step 1: Prepare data
# Convert all categorical variables to dummy variables
# Step 1: Prepare data
lasso_data <- rsf_data %>%
  select(survival_months, event, mets_lung, mets_liver, mets_bone, mets_brain,
         age_group, race, median_income, diagnosis_year) %>%
  filter(if_all(everything(), ~ !is.na(.))) %>%      # Remove rows with any NA
  filter(survival_months > 0)                        # Remove 0 or negative times

# Create model matrix (dummy coding all predictors)
x <- model.matrix(~ mets_lung + mets_liver + mets_bone + mets_brain +
                    age_group + race + median_income + diagnosis_year,
                  data = lasso_data)[, -1]  # drop intercept

# Create survival object
y <- Surv(lasso_data$survival_months, lasso_data$event)

# Step 2: Run cross-validated LASSO Cox
set.seed(123)
cvfit <- cv.glmnet(x, y, family = "cox", alpha = 1)

# Plot cross-validation curve
plot(cvfit)

# Step 3: Extract best lambda and coefficients
best_lambda <- cvfit$lambda.min
lasso_coef <- coef(cvfit, s = "lambda.min")

# Convert to tidy format
lasso_selected <- as.data.frame(as.matrix(lasso_coef))
lasso_selected <- lasso_selected %>%
  tibble::rownames_to_column("variable") %>%
  rename(coefficient = `1`) %>%
  filter(coefficient != 0)

# Step 4: View selected variables
print(lasso_selected)

# Optional: Save selected variables
write.csv(lasso_selected, "results/tables/lasso_selected_variables.csv", row.names = FALSE)

```{r}
####################################################
# LASSO Cox: AUC and Confusion Matrix at 12 Months
####################################################

# Step 1: Use same cleaned dataset
# Make sure this matches the data used to train LASSO
lasso_data <- rsf_data %>%
  select(survival_months, event, mets_lung, mets_liver, mets_bone, mets_brain,
         age_group, race, median_income, diagnosis_year) %>%
  filter(if_all(everything(), ~ !is.na(.))) %>%
  filter(survival_months > 0)

# Model matrix
x <- model.matrix(~ mets_lung + mets_liver + mets_bone + mets_brain +
                    age_group + race + median_income + diagnosis_year,
                  data = lasso_data)[, -1]

# Step 2: Predict linear risk scores
risk_scores <- predict(cvfit, newx = x, s = "lambda.min", type = "link")

# Step 3: Define binary outcome at 12 months
target_time <- 12
true_class <- ifelse(lasso_data$survival_months <= target_time & lasso_data$event == 1, 1, 0)

# Step 4: AUC
roc_obj <- roc(response = true_class, predictor = as.vector(risk_scores))
auc_val <- auc(roc_obj)
print(paste("AUC at 12 months:", round(auc_val, 3)))

# Step 5: Confusion Matrix (You can choose threshold = 0 by default or use Youden index)
predicted_class <- ifelse(risk_scores > 0, 1, 0)

# Confusion Matrix
conf_matrix <- confusionMatrix(factor(predicted_class), factor(true_class), positive = "1")
print(conf_matrix)

## Comparing the models

I compared three modeling approaches to predict 12-month survival outcomes: Random Survival Forest (RSF), LASSO Cox regression, and traditional Random Forest (RF). RSF achieved the highest AUC (0.74), indicating superior discrimination, and balanced sensitivity (86.5%) and specificity (47.9%). LASSO Cox achieved slightly lower AUC (0.66) but had comparable sensitivity and slightly better accuracy. The RF classifier performed poorly (AUC = 0.39), highlighting the limitations of using classification models for censored survival data. RSF is optimal for prediction, while LASSO offers interpretability and variable selection.

```{r}
####################################################
# Compare RSF vs LASSO Cox: AUC & Confusion Matrix
# Fixed time point: 12 months
####################################################

# Step 1: Common dataset for both models
eval_data <- rsf_data %>%
  select(survival_months, event, mets_lung, mets_liver, mets_bone, mets_brain,
         age_group, race, median_income, diagnosis_year) %>%
  filter(if_all(everything(), ~ !is.na(.))) %>%
  filter(survival_months > 0)

# Define event status by 12 months
target_time <- 12
true_class <- ifelse(eval_data$survival_months <= target_time & eval_data$event == 1, 1, 0)

##############################################
# Part 1: RSF Model Predictions
##############################################
# Predict survival probabilities
rsf_preds <- predict(rsf_model, data = eval_data, type = "response")
rsf_surv12 <- rsf_preds$survival[, which.min(abs(rsf_model$unique.death.times - target_time))]
rsf_pred_class <- ifelse(rsf_surv12 < 0.5, 1, 0)

# AUC
rsf_roc <- roc(response = true_class, predictor = 1 - rsf_surv12)
rsf_auc <- auc(rsf_roc)
print(paste("RSF AUC:", round(rsf_auc, 3)))

# Confusion Matrix
rsf_cm <- confusionMatrix(factor(rsf_pred_class), factor(true_class), positive = "1")
print(rsf_cm)

##############################################
# Part 2: LASSO Cox Model Predictions
##############################################
# Create model matrix
x_lasso <- model.matrix(~ mets_lung + mets_liver + mets_bone + mets_brain +
                          age_group + race + median_income + diagnosis_year,
                        data = eval_data)[, -1]

# Linear risk scores from LASSO Cox
lasso_scores <- predict(cvfit, newx = x_lasso, s = "lambda.min", type = "link")
lasso_pred_class <- ifelse(lasso_scores > 0, 1, 0)

# AUC
lasso_roc <- roc(response = true_class, predictor = as.vector(lasso_scores))
lasso_auc <- auc(lasso_roc)
print(paste("LASSO Cox AUC:", round(lasso_auc, 3)))

# Confusion Matrix
lasso_cm <- confusionMatrix(factor(lasso_pred_class), factor(true_class), positive = "1")
print(lasso_cm)

### 2025-04-03 products/manuscript/Manuscript.qmd
part 4 project

@@

## Basic statistical analysis
### Bivariate Analysis: Metastasis and Survival

@@

```{r}
#| label: fig-km-curve-lung-mets
#| fig-cap: "Kaplan-Meier survival curves for Lung Metastasis"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "KM_curve_lung_mets.png"))
```
### Bivariate Analysis: Race and Survival
Patients identified as  had significantly worse survival outcomes than White patients, with a 37% higher risk of death (HR = 1.37, p < 0.0001). In contrast, Asian patients had a lower risk of death compared to White patients (HR = 0.86, p = 0.0076), suggesting a potential survival advantage. The survival difference between American Indian and White patients was not statistically significant (HR = 1.21, p = 0.29), indicating no clear evidence of an increased or decreased risk. @tbl-summaryhrrace shows the HR for patients with different race.
```{r}
#| label: tbl-summaryhrrace
#| tbl-cap: "Proportional hazards analysis for different Race."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "race_HR.rds"))
knitr::kable(resulttable)
```
### Bivariate Analysis: Age and Survival
The hazard ratio (HR) estimates for different age groups indicate a general trend of increasing risk with age, though statistical significance varies. Compared to the reference group (20-24 years), younger age groups (25-74 years) show HRs between 3.5 and 6.2, but their p-values are above 0.05, suggesting no statistically significant difference in survival risk. However, individuals aged 80-84 years (HR = 7.99, p = 0.038) and 85+ years (HR = 11.01, p = 0.017) have significantly higher mortality risk, indicating poorer survival outcomes in older patients. These findings suggest that age is an important predictor of survival, with the highest risk observed in the oldest age group. @tbl-summaryhrage shows the HR for diferrent age groups.
```{r}
#| label: tbl-summaryhrage
#| tbl-cap: "Proportional hazards analysis for different age group."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "age_HR.rds"))
knitr::kable(resulttable)
```
### Multivariable analysis

The multivariable Cox proportional hazards model results indicate that metastasis to the bone (HR = 1.30, p < 0.001), brain (HR = 1.72, p < 0.001), liver (HR = 1.66, p < 0.001), and lung (HR = 1.51, p < 0.001) are all significantly associated with poorer survival outcomes. Among racial groups, Black patients (HR = 1.39, p < 0.001) had significantly worse survival compared to White patients, while Asian or Pacific Islander and American Indian/Alaska Native groups did not show a statistically significant difference.

@@

Year of diagnosis did not show a consistent or significant trend, suggesting no clear improvement or decline in survival based on the year of diagnosis. Household income was associated with survival, with patients in the $50,000 - $54,999 (HR = 1.16, p = 0.011) and $65,000 - $69,999 (HR = 1.12, p = 0.030) categories showing significantly higher risk compared to those earning $75,000+. However, other income groups did not show significant differences.

@@

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

>>>>

### Machine leaning model

I applied a Random Survival Forest (RSF) model to evaluate the impact of clinical and demographic variables on overall survival time.
RSF is a non-parametric ensemble learning method specifically designed for right-censored survival data.
It extends the traditional random forest approach by using survival splitting rules and ensemble cumulative hazard estimates.
The outcome of interest was overall survival time in months, with vital status (0 = alive, 1 = dead) as the event indicator.
Predictor variables included metastasis status (bone, liver, lung, brain), age group, race, year of diagnosis, and median household income.
All categorical variables were converted to factors, as required by the randomForestSRC package in R.
The model was trained using 500 trees, and variable importance was assessed using the impurity measure.
Model performance was evaluated through time-dependent ROC analysis, confusion matrix, and individualized survival curve prediction.
The RSF approach enabled flexible modeling of complex interactions and non-linear effects, offering an interpretable and robust alternative to traditional Cox models.

@@

## Basic statistical analysis

### Bivariate Analysis: Metastasis and Survival

@@

```{r}
#| label: fig-km-curve-lung-mets
#| fig-cap: "Kaplan-Meier survival curves for Lung Metastasis"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "KM_curve_lung_mets.png"))
```

### Bivariate Analysis: Race and Survival

Patients identified as had significantly worse survival outcomes than White patients, with a 37% higher risk of death (HR = 1.37, p \< 0.0001).
In contrast, Asian patients had a lower risk of death compared to White patients (HR = 0.86, p = 0.0076), suggesting a potential survival advantage.
The survival difference between American Indian and White patients was not statistically significant (HR = 1.21, p = 0.29), indicating no clear evidence of an increased or decreased risk.
@tbl-summaryhrrace shows the HR for patients with different race.

```{r}
#| label: tbl-summaryhrrace
#| tbl-cap: "Proportional hazards analysis for different Race."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "race_HR.rds"))
knitr::kable(resulttable)
```

### Bivariate Analysis: Age and Survival

The hazard ratio (HR) estimates for different age groups indicate a general trend of increasing risk with age, though statistical significance varies.
Compared to the reference group (20-24 years), younger age groups (25-74 years) show HRs between 3.5 and 6.2, but their p-values are above 0.05, suggesting no statistically significant difference in survival risk.
However, individuals aged 80-84 years (HR = 7.99, p = 0.038) and 85+ years (HR = 11.01, p = 0.017) have significantly higher mortality risk, indicating poorer survival outcomes in older patients.
These findings suggest that age is an important predictor of survival, with the highest risk observed in the oldest age group.
@tbl-summaryhrage shows the HR for diferrent age groups.

```{r}
#| label: tbl-summaryhrage
#| tbl-cap: "Proportional hazards analysis for different age group."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "age_HR.rds"))
knitr::kable(resulttable)
```

### Multivariable analysis

The multivariable Cox proportional hazards model results indicate that metastasis to the bone (HR = 1.30, p \< 0.001), brain (HR = 1.72, p \< 0.001), liver (HR = 1.66, p \< 0.001), and lung (HR = 1.51, p \< 0.001) are all significantly associated with poorer survival outcomes.
Among racial groups, Black patients (HR = 1.39, p \< 0.001) had significantly worse survival compared to White patients, while Asian or Pacific Islander and American Indian/Alaska Native groups did not show a statistically significant difference.

@@

Year of diagnosis did not show a consistent or significant trend, suggesting no clear improvement or decline in survival based on the year of diagnosis.
Household income was associated with survival, with patients in the \$50,000 - \$54,999 (HR = 1.16, p = 0.011) and \$65,000 - \$69,999 (HR = 1.12, p = 0.030) categories showing significantly higher risk compared to those earning \$75,000+.
However, other income groups did not show significant differences.

@@

### Random Survival Forest

A Random Survival Forest (RSF) model was used to evaluate the contribution of clinical and demographic predictors to overall survival.
The RSF model demonstrated strong performance in predicting survival at 12 months, achieving an AUC of 0.74 (@fig-rsf-auc).
The confusion matrix revealed a sensitivity of 86.5% and a specificity of 47.9%, with an overall accuracy of 53.8% and balanced accuracy of 67.2%, indicating that the model was effective in identifying patients at high risk of death within the first year.

Variable importance analysis revealed that age group, median household income, and year of diagnosis were the most influential predictors of survival (@fig-rsf-variable-importance).
Other important variables included race and metastasis to the bone, lung, liver, and brain.
The individualized survival prediction for a representative patient showed a steady decline in survival probability over time, reflecting the nature of long-term risk in this cohort (@fig-rsf-ind-sur-pre). @fig-rsf-con-matrix. Confusion matrix showing prediction results of the Random Survival Forest model at 12 months. The model correctly classified 709 survivors and 2,182 non-survivors, with moderate misclassification of 2,376 survivors and 111 non-survivors.

```{r}
#| label: fig-rsf-auc
#| fig-cap: "Random Survival Forest AUC"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_auc_12mo.png"))
```

```{r}
#| label: fig-rsf-variable-importance
#| fig-cap: "Random Survival Forest Variable Importance"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_variable_importance.png"))
```
```{r}
#| label: fig-rsf-ind-sur-pre
#| fig-cap: "Random Survival Forest Individualized Survival Prediction"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_survival_curve_patient1.png"))
```
```{r}
#| label: fig-rsf-con-matrix
#| fig-cap: "RSF Confusion Matrix (12 Months)"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_conf_matrix_12mo.png"))
```
{{< pagebreak >}}
